Trilemma

### 2022-02-08 Trilemma_clustering.Rmd
back up trilemma clustering analysis

---
title: "Trilemma"
author: "Juan Rocha"
date: "Updated `r Sys.Date()`"
output:
  html_notebook:
    toc: yes
    toc_float: yes
    highlight: tango
    code_folding: hide
    df_print: paged
    theme: 
      bootswatch: cosmo
      code_font:
        google: Fira Code

---

@@

# theme_set(theme_light(base_size = 10))
```

@@

```{r}
dat1 <- read_csv(
    "~/Documents/Projects/DATA/WorldBank/SDG_csv/SDGData.csv") %>% 
    janitor::clean_names()
```

@@

dat <- dat %>% 
  left_join(
    ., 
    dat1 %>% 
      select(-x33) %>% 
      filter(indicator_name == "GNI per capita (current US$)") %>% 
      pivot_longer(cols = x1990:x2017, names_to = "year", values_to = "value") %>%
      mutate(year = str_remove(year, "x")) %>% 
      mutate(year = as.numeric(year)) %>% 
      mutate(ismissing = is.na(value)) %>% 
      filter(year >= 1990) %>% 
      rename(country = country_name, gni = value)
  )

@@

```{r message=FALSE, warning=FALSE}
df_dat %>% 
    select(year, EFconsPerCap, gni2, gini) %>% 
    GGally::ggpairs(
        columns = 3:5 
    # lower = list(continuous = GGally::wrap("points", alpha = 0.5, size = 0.5)),
    # diag = list(continuous = GGally::wrap("densityDiag", alpha = 0.2)),
    # upper = list(continuous = GGally::wrap("cor", size = 2))
    ) + theme_light(base_size = 16)
```

@@

p1 + p2 + p3 + plot_layout(guides = "collect") & theme_light(base_size = 15)
```

```{r}
df_dat %>% 
    as_tibble() %>% 
    group_by(country) %>% 
    summarize(
        mean_ef = mean(EFconsPerCap),
        mean_gni = mean(gni2),
        mean_gini = mean(gini)
    ) %>% 
    arrange(desc(mean_ef))
```
### Experimenting with phase plane

```{r}
df_dat |> 
    mutate(
        d_EF = slide_dbl(EFconsPerCap, diff, .before = 1, .after = 0, .complete = TRUE),
        d_gini = slide_dbl(gini, diff, .before = 1, .after = 0, .complete = TRUE),
        d_gni2 = slide_dbl(gni2, diff, .before = 1, .after = 0, .complete = TRUE)
    ) |> 
    mutate(
        theta = atan(d_gini/d_EF),
        gamma = atan(d_gini/d_gni2)
    )
```

@@

```{r}
m <- "ward.D2" # minimize the total within-cluster variance

clust_num <- NbClust::NbClust( 
    data = df_dat %>% as.data.frame() %>% 
        select(where(is.numeric), -year),
    #distance = "maximum",
    min.nc = 2, max.nc = 10, 
    method = m, 
    index = 'all')
```

@@

```{r}
## Stability & Internal validation
stab <- clValid::clValid(
    obj = df_dat %>% as.data.frame() %>%
        select(where(is.numeric), -year) %>%
        as.matrix(),
    nClust = c(2:9),
    clMethods = c("hierarchical", "kmeans",  #"som", "model", "diana", "fanny",
                 "sota", "pam", "clara", "agnes"),
    validation = c('stability', "internal"),
    #metric = "manhattan",
    method = "ward",
    verbose = FALSE
) ## Hierarchical is the optimal method

@@

>>>>

---
title: "Trilemma"
author: "Juan Rocha"
date: "Updated `r Sys.Date()`"
output:
  html_notebook:
    toc: yes
    toc_float: yes
    highlight: tango
    code_folding: hide
    df_print: paged
    theme: 
      bootswatch: cosmo
      code_font:
        google: Fira Code
editor_options: 
  chunk_output_type: inline
---

@@

#theme_set(theme_light(base_size = 12))
```

@@

```{r}
dat1 <- read_csv(
    "~/Documents/Projects/DATA/WorldBank/SDG_csv/SDGData.csv") %>% 
    janitor::clean_names()

missingness <- dat1 %>% 
    select(-last_col()) %>% 
    filter(str_detect(indicator_name, "GNI")) %>% 
    pivot_longer(cols = x1990:x2019, names_to = "year", values_to = "value") %>% 
    mutate(year = str_remove(year, "x")) %>% 
    mutate(year = as.numeric(year)) %>% 
    mutate(ismissing = is.na(value)) %>% 
    group_by(country_code, indicator_name, country_name) %>% 
    summarize(missingness = sum(ismissing)/n())

missingness%>% 
    ggplot(aes(country_code, indicator_name)) +
    geom_tile(aes(fill = missingness)) + 
    scale_fill_viridis_c(option = "C") +
    theme(axis.text.x = element_text(angle =90), legend.position = "top")
```

@@

dat <- dat %>% 
  left_join(
    ., 
    dat1 %>% 
      select(-last_col()) %>% 
      filter(indicator_name == "GNI per capita (US$)") %>% 
      pivot_longer(cols = x1990:x2019, names_to = "year", values_to = "value") %>%
      mutate(year = str_remove(year, "x")) %>% 
      mutate(year = as.numeric(year)) %>% 
      mutate(ismissing = is.na(value)) %>% 
      filter(year >= 1990) %>% 
      rename(country = country_name, gni = value)
  )

@@

```{r message=FALSE, warning=FALSE}
df_dat %>% 
    select(year, EFconsPerCap, gni2, gini) %>% 
    GGally::ggpairs(
        columns = 3:5 
    # lower = list(continuous = GGally::wrap("points", alpha = 0.5, size = 0.5)),
    # diag = list(continuous = GGally::wrap("densityDiag", alpha = 0.2)),
    # upper = list(continuous = GGally::wrap("cor", size = 2))
    ) + theme_light(base_size = 20)
```

@@

p1 + p2 + p3 + plot_layout(guides = "collect") & theme_light(base_size = 18)
```

```{r}
df_dat %>% 
    as_tibble() %>% 
    group_by(country) %>% 
    summarize(
        mean_ef = mean(EFconsPerCap),
        mean_gni = mean(gni2),
        mean_gini = mean(gini),
        var_ef = var(EFconsPerCap),
        var_gni = var(gni2),
        var_gini = var(gini)
    ) %>% 
    arrange(desc(mean_ef))
```
### Experimenting with phase plane

```{r}
df_dat <- df_dat |> 
    # change scale of GNI to log so the magnitud is not so dominated by $
    mutate(gni_log = log(gni2)) |> 
    mutate(
        d_EF = slide_dbl(EFconsPerCap, diff, .before = 1, .after = 0, .complete = TRUE),
        d_gini = slide_dbl(gini, diff, .before = 1, .after = 0, .complete = TRUE),
        d_gni = slide_dbl(gni_log, diff, .before = 1, .after = 0, .complete = TRUE)
    ) |> 
    mutate(
        magni = sqrt(d_EF^2 + d_gini^2 + d_gni^2), 
        theta = atan(d_gini/d_EF),
        gamma = atan(d_gini/d_gni)
    ) 
```

```{r}
df_dat |> 
    as_tibble() |> 
    ungroup() |> 
    group_by(country) |> 
    summarize(mgt = sum(magni, na.rm = TRUE), 
              theta = sum(theta, na.rm = TRUE),
              gamma = sum(gamma, na.rm = TRUE)) |> 
    arrange(desc(mgt))
```
```{r}
df_dat |> 
    as_tibble() |> 
    select(country, year, d_EF, d_gini, d_gni) |> 
    pivot_longer(cols = starts_with("d_"), names_to = "vars", values_to = "vals") |> 
    filter(!is.na(vals)) |> 
    ggplot(aes(year, vals, group = country)) +
    geom_path(size = 0.2, alpha = 0.5) +
    facet_wrap(~vars, scales = "free") +
    theme_light(base_size = 18)
```
```{r}
df_dat |> 
    ggplot(aes(EFconsPerCap, gini)) +
    geom_density2d() +
    geom_point(aes(color = gni_log)) +
    geom_path(aes(group = country), size = 0.15) +
    scico::scale_color_scico() +
    theme_light(base_size = 18)
```
```{r}
df_dat |> 
    ggplot(aes(gni_log, gini)) +
    geom_density2d() +
    geom_point(aes(color = EFconsPerCap)) +
    geom_path(aes(group = country), size = 0.15) +
    scico::scale_color_scico(palette = "berlin") +
    theme_light(base_size = 18)
```
```{r}
df_dat |> 
    ggplot(aes(gni_log, EFconsPerCap)) +
    geom_density2d() +
    geom_point(aes(color = gini)) +
    geom_path(aes(group = country), size = 0.15) +
    scico::scale_color_scico() +
    theme_light(base_size = 18)
```
Who is the weird trajectory out there, world? No it's Luxemburg

```{r}
df_dat |> 
    arrange(desc(EFconsPerCap))
```

@@

```{r}
df_summary <- df_dat |> 
    as_tibble() |> 
    ungroup() |> 
    group_by(country) |> 
    summarize(m = sum(magni, na.rm = TRUE), 
              t = sum(theta, na.rm = TRUE),
              g = sum(gamma, na.rm = TRUE),
              sd_mag = sd(magni, na.rm = TRUE),
              sd_t = sd(theta, na.rm = TRUE),
              sd_g = sd(gamma, na.rm = TRUE), 
              ef = mean(EFconsPerCap, na.rm = TRUE),
              inq = mean(gini, na.rm = TRUE),
              gni = mean(gni_log, na.rm = TRUE),
              #sd_ef = sd(ef, na.rm = TRUE),
              sd_inq = sd(gini, na.rm = TRUE),
              sd_gni = sd(gni_log, na.rm = TRUE))

m <- "ward.D2" # minimize the total within-cluster variance

clust_num <- NbClust::NbClust( 
    data = df_summary %>% as.data.frame() %>% 
        select(where(is.numeric)),
    #distance = "maximum",
    min.nc = 2, max.nc = 7, 
    method = m, 
    index = 'all')
```

@@

```{r}
## Stability & Internal validation
stab <- clValid::clValid(
    obj = df_summary %>% as.data.frame() %>% 
        select(where(is.numeric)) %>%
        as.matrix(),
    nClust = c(2:7),
    clMethods = c("hierarchical", "kmeans",  #"som", "model", "diana", "fanny",
                 "sota", "pam", "clara", "agnes"),
    validation = c('stability', "internal"),
    #metric = "manhattan",
    method = "ward",
    verbose = FALSE
) ## Hierarchical is the optimal method

@@

```{r}
stab@clusterObjs$pam$`2`$clustering
```

```{r fig.width=7, fig.height=8}
df_summary |> 
    mutate(clus = stab@clusterObjs$pam$`2`$clustering) |> 
    arrange(desc(m)) |> 
    mutate(country = as_factor(country)) |> 
    mutate(country = fct_reorder2(country, m,g)) |> 
    pivot_longer(cols = m:sd_gni, names_to = "vars", values_to = "vals") |> 
    ggplot(aes(vals, country)) +
    geom_col(aes(fill = as.factor(clus))) +
    facet_grid(clus~vars, scales = "free") +
    theme_light(base_size = 10)
```

```{r}
df_dat |> 
    as_tibble() |> 
    group_by(country) |> 
    summarize(
        ef = sum(d_EF, na.rm = TRUE),
        gini = sum(d_gini, na.rm = TRUE),
        gni = sum(d_gni, na.rm = TRUE)
    ) |> 
    mutate(clus = stab@clusterObjs$pam$`2`$clustering) |> 
    ggplot(aes(ef, gini)) +
    geom_segment(
        aes(x = 0, y = 0, xend = ef, yend = gini, color = as.factor(clus)), arrow = arrow(length = unit(0.5, "cm"))) +
    geom_text(aes(label = country)) +
    theme_light(base_size = 20)
```

```{r}
df_dat |> 
    as_tibble() |> 
    group_by(country) |> 
    summarize(
        ef = sum(d_EF, na.rm = TRUE),
        gini = sum(d_gini, na.rm = TRUE),
        gni = sum(d_gni, na.rm = TRUE)
    ) |> 
    mutate(clus = stab@clusterObjs$pam$`2`$clustering) |> 
    ggplot(aes(gni, gini)) +
    geom_segment(aes(x = 0, y = 0, xend = gni, yend = gini, color = as.factor(clus)), arrow = arrow(length = unit(0.5, "cm"))) +
    geom_text(aes(label = country)) +
    theme_light(base_size = 20)
```

### Old stuff
